---
title: "Answers to Exercises Day 1"
author: "Aditya Gudal, adapted from exercises by Christina Maimone"
date: "`r Sys.Date()`"
output: html_document
params:
    answers: TRUE
---


```{r, echo=FALSE, eval=TRUE}
answers<-params$answers
```

```{r global_options, echo = FALSE, include = FALSE}
knitr::opts_chunk$set(echo=answers, eval=answers,
                      warning = FALSE, message = FALSE,
                      cache = FALSE, tidy = FALSE)
```

## Load the data

Load the `gapminder` dataset.

```{r 1,echo=T,include=T}
gapminder <- read.csv(here::here("data/gapminder5.csv"), stringsAsFactors=FALSE)
str(gapminder)
```


## If Statement

Use an if() statement to print a suitable message reporting whether there are any records from 2002 in the gapminder dataset. Now do the same for 2012.

Hint: use the `any` function.

```{r 2,echo=T,include=T}
year<-2002
if(any(gapminder$year==year)){
  print('yes')
}else{
 print('no') 
}
```



## Loop and If Statements

Write a script that finds the mean life expectancy by country for countries whose population is below the mean for the dataset

Write a script that loops through the `gapminder` data by continent and prints out whether the mean life expectancy is smaller than 50, between 50 and 70, or greater than 70.

```{r ans1,echo=T,include=T}
pop_mean<-mean(gapminder$pop)
for(i in unique(gapminder$country)){
  country_mean<-mean(gapminder$pop[gapminder$country==i])
  
  if(country_mean<pop_mean){
    print(paste('the mean life expectancy of ',i,'is',mean(gapminder$lifeExp[gapminder$country==i])))
  }
}
```

```{r ans2,echo=T}
l_thresh<-50
h_thresh<-70

for (continent in unique(gapminder$continent)){
  meanlifeExp<-mean(gapminder$lifeExp[gapminder$continent==continent])
  
  if (meanlifeExp<l_thresh){
    print(paste(continent,'has low life Expectancy'))
  }
  else if(meanlifeExp  %in% c(l_thresh,h_thresh)){
    print(paste(continent,'has medium life expectancy'))
  }
  else{
    print(paste(continent,'has high life expectancy'))
  }
}
```
## Exercise: Write Functions

Create a function that given a data frame will print the name of each column and the class of data it contains. Use the gapminder dataset.  Hint: Use `mode()` or `class()` to get the class of the data in each column. Remember that `names()` or `colnames()` returns the name of the columns in a dataset.

```{r creating func,echo=T}
name_and_type<- function(data) {
  cols <- names(data)
  for (i in cols) {
    print(paste0(i, ": ", class(data[, i])))
    }
}
name_and_type(gapminder)

#Note: Some of these were taken or modified from #https://www.r-bloggers.com/functions-exercises/
```
Create a function that given a vector will print the mean and the standard deviation of a **vector**, it will optionally also print the median. Hint: include an argument that takes a boolean (`TRUE`/`FALSE`) operator and then include an `if` statement.

```{r creating function,include=T,echo=T}
vector_given<-function(x,include_median=F){
  print(paste('Mean : ',mean(x)))
  print(paste('Standard Deviation : ',sd(x)))
  if(include_median){
    print(paste('Median : ',median(x)))
  }
}

vector_given(gapminder$year,T)

```

## Analyzing the relationship 

Use what you've learned so far to answer the following questions using the `gapminder` dataset. Be sure to include some visualizations!

1. What is the relationship between GDP per capita and life expectancy? Does this relationship change over time? (Hint: Use the natural log of both variables.)
```{r ques1,include=T,echo=T}
years<-unique(gapminder$year)

data_1<-data.frame(x=years,y=sapply(years,function(x) cor(log(gapminder$gdpPercap[x==gapminder$year]),log(gapminder$lifeExp[x==gapminder$year]))))

fit <- lm(data_1$y ~data_1$x)

summary(fit)
```


```{r ques1 plot,echo=T}
plot(data_1$y ~ data_1$x, main ='Relationship between GDP per capita and Life Expectancy per year',xlab='Year',ylab='Correlated Relationship')
lines(loess.smooth(data_1$x, data_1$y,col='red',degree=10))



```



2. Does the relationship between GDP per capita and life expectacy vary by continent? Make sure you divide the Americas into North and South America.

```{r ques2 relationship,echo=T}
gapminder$continent[gapminder$continent=='Americas']<-ifelse(gapminder$country%in%c('Canada','Costa Rica','Cuba','Dominican Republic','El Salvador','Guatemala','Haiti','Honduras','Jamaica','Mexico','Nicaragua','Panama','Puerto Rico','Trinidad and Tobago','United States'),'North America','South America')

unique_continents<-unique(gapminder$continent)

data_2<-data.frame(y=sapply(unique_continents,function(x) cor(log(gapminder$gdpPercap[x==gapminder$continent]),log(gapminder$lifeExp[x==gapminder$continent]))))

data_2
```